Owlifier: Creating OWL-DL Ontologies from Simple Spreadsheet-Based Knowledge Descriptions\tnoteref{t1}

### 2008-12-03 docs/pubs/ISEI6/main.tex

\documentclass[preprint,number]{elsarticle} 
% other options: preprint, review, 1p, 3p, 5p, 
% bib options: authoryear, number, longtitle,
% other options: times

@@

The rest of this paper is organized as follows. In
\secref{sec:owlifier} we describe the basic syntax and semantics of
\owlifier. We define blocks that support a large subset of OWL-DL and
that also generally follow the ontology creation guidelines defined in
\cite{rector04:_owl_pizzas}. We also simplify certain aspects of
ontology creation using OWL-DL, e.g., by assuming classes are disjoint
by default (unless specified otherwise) and by applying implicit
property restriction closures \cite{rector04:_owl_pizzas}. In
\secref{sec:characteristics} we describe additional characteristics of
\owlifier\ and discuss issues with respect to classification and
reasoning. In \secref{sec:implementation} we briefly describe the
\owlifier\ implementation, and conlcude in \secref{sec:conclusion}
with related and future work. In general, the goal of \owlifier\ is
not to support all constructs in OWL-DL, but instead to provide a
higher-level ontology syntax (via spreadsheet blocks) that is easy for
ecologists to use and understand while also providing the necessary
constructs for developing typical ecological ontologies. By compiling
\owlifier\ to OWL-DL, we also allow for experts to refine and extend
the ontology using more advanced ontology editing tools if necessary.

@@

\myblock{Overlap Blocks.} Except in certain situations (described
further in \secref{sec:characteristics}), classes are generally
assumed to be disjoint in \owlifier.  Overlap blocks explicitly relax
this assumption by stating that a given set of classes may have
overlapping instances. Rows containing overlap blocks take the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \textsf{overlap} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where each class $c_i$ is allowed to share instances with each class
$c_j$, for $1 \le i,j \le n$. That is, a given $c_i$ and $c_j$ in an
overlap block are not defined to be disjoint classes in the current
ontology. As an example, consider the following entity blocks that
define the classes Estuary, Lagoon, and Marsh as subclasses of
Ecological Habitats.
\begin{tabbing}
  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Estuary} \\ 
  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Lagoon} \\ 
  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Marsh} 
\end{tabbing}
Given only these blocks, \owlifier\ treats Estuary, Lagoon, and Marsh
as disjoint classes. To relax this assumption and allow, e.g., types
of Lagoons to also be types of Estuaries, we explicitly add the
following overlap block
\begin{tabbing}
  ~~\textsf{overlap} \textsf{Estuary} \textsf{Lagoon}
\end{tabbing}
In general, overlap blocks are rarely used but provide a mechanism to
override the default behavior of \owlifier\ in asserting disjoint
classes.

@@

\section{Additional Features of \Owlifier}
\label{sec:characteristics}

In this section, we briefly describe some of the additional features
of \owlifier, specifically focusing on the use of disjoint classes,
property closures, \owlifier\ reasoning, and additional block syntax.

@@

OWL is based on the open world assumption, which can lead to a number
of ontology development ``pitfalls'' for those new to the language
\cite{smith04:_owl_web_ontol_languag_guide,rector04:_owl_pizzas}. One
example is in the creation of disjoint classes (another concerns
property closures, described below). In particular, unless explicitly
stated, distinct classes within an OWL-DL ontology are never assumed
to be disjoint. However, in many ontologies a large number of classes
are typically defined as being disjoint (e.g., sibling classes), and
stating these disjoint constraints is often time-consuming since each
pair of classes must be given an explicit disjoint assertion. Editors
such as Protege \cite{knublauch04:_editin_descr_logic_ontol_with}
provide shortcuts via the user interface to create specific sets of
disjoint assertions, e.g., by allowing a user to define all children
of a particular class as disjoint. In general, however, many users
expect such classes to be disjoint by default
\cite{rector04:_owl_pizzas} and this expectation often leads to
modeling errors.

@@

\subsection{Object Property Closure}

Another common mistake for new users of OWL-DL concerns the user of
universal rather than existential property restrictions
\cite{rector04:_owl_pizzas}. In particular, an expression such as
$\texttt{Body} \sqsubseteq \forall \textsf{hasPart}$

\subsection{Reasoning in \owlifier}

4. is reasonable, e.g., it is possible to define values, value types,
and untangling

@@

1. non-ambiguous blocks (no block leads to an ambiguous DL axiom)

5. we allow sparse blocks, and blocks can be given in any order

\section{Implementation of \Owlifier}
\label{sec:implementation}

figure showing basics owlifier interaction, description.

current status of implementation (subset of the syntax described here), perform classification via pellet/jena, performs disjoint check

ongoing extensions: additional syntax; specify delimeter characters;
perform inverse translations (easy to perform, e.g., ...); perform
closure computation (again, easy).

@@

related work: similar to dave's work on ``global taxonomic
constraints'', which assume certain structures of hierarchies, e.g.,
disjoint siblings. protege approach/plugin

>>>>

\documentclass[preprint,number]{elsarticle} 
% other options: preprint, review, 1p, 3p, 5p, 
% bib options: authoryear, number, longtitle,
% other options: times
\usepackage{graphicx}

@@

The rest of this paper is organized as follows. In
\secref{sec:owlifier} we describe the basic syntax and semantics of
\owlifier. We define blocks that support a large subset of OWL-DL and
that also generally follow the ontology creation guidelines defined in
\cite{rector04:_owl_pizzas}. We also simplify certain aspects of
ontology creation using OWL-DL, e.g., by assuming classes are disjoint
by default (unless specified otherwise) and supporting sufficient
restrictions \cite{rector04:_owl_pizzas}. In
\secref{sec:features} we describe additional characteristics of
\owlifier\ and discuss issues with respect to classification and
reasoning. In \secref{sec:implementation} we briefly describe the
\owlifier\ implementation, and conlcude in \secref{sec:conclusion}
with related and future work. In general, the goal of \owlifier\ is
not to support all constructs in OWL-DL, but instead to provide a
higher-level ontology syntax (via spreadsheet blocks) that is easy for
ecologists to use and understand while also providing the necessary
constructs for developing typical ecological ontologies. By compiling
\owlifier\ to OWL-DL, we also allow for experts to refine and extend
the ontology using more advanced ontology editing tools if necessary.

@@

\myblock{Overlap Blocks.} Except in certain situations (described
further in \secref{sec:features}), classes are generally
assumed to be disjoint in \owlifier.  Overlap blocks explicitly relax
this assumption by stating that a given set of classes may have
overlapping instances. Rows containing overlap blocks take the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \textsf{overlap} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where each class $c_i$ is allowed to share instances with each class
$c_j$, for $1 \le i,j \le n$. That is, a given $c_i$ and $c_j$ in an
overlap block are not defined to be disjoint classes in the current
ontology. As an example, consider the following entity blocks that
define the classes Estuary, Lagoon, and Marsh as subclasses of
Ecological Habitats.
\begin{tabbing}
  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Estuary} \\ 
  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Lagoon} \\ 
  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Marsh} 
\end{tabbing}
Given only these blocks, \owlifier\ treats Estuary, Lagoon, and Marsh
as disjoint classes. To relax this assumption and allow, e.g., types
of Lagoons to also be types of Estuaries, we explicitly add the
following overlap block
\begin{tabbing}
  ~~\textsf{overlap} \textsf{Estuary} \textsf{Lagoon}
\end{tabbing}
In general, overlap blocks are rarely used but provide a mechanism to
override the default behavior of \owlifier\ in asserting disjoint
classes.

@@

\section{Additional Features of \Owlifier}
\label{sec:features}

In this section, we briefly describe some of the additional features
of \owlifier, specifically focusing on the use of disjoint classes,
\owlifier\ reasoning, and additional block syntax.

@@

OWL is based on the open world assumption, which can lead to a number
of ontology development ``pitfalls'' for those new to the language
\cite{smith04:_owl_web_ontol_languag_guide,rector04:_owl_pizzas}. One
example is in the creation of disjoint classes.  In particular, unless
explicitly stated, distinct classes within an OWL-DL ontology are
never assumed to be disjoint. However, in many ontologies a large
number of classes are typically defined as being disjoint (e.g.,
sibling classes), and stating these disjoint constraints is often
time-consuming since each pair of classes must be given an explicit
disjoint assertion. Editors such as Protege
\cite{knublauch04:_editin_descr_logic_ontol_with} provide shortcuts
via the user interface to create specific sets of disjoint assertions,
e.g., by allowing a user to define all children of a particular class
as disjoint. In general, however, many users expect such classes to be
disjoint by default \cite{rector04:_owl_pizzas} and this expectation
often leads to modeling errors.

@@

\subsection{Reasoning in \Owlifier}

Blocks in \owlifier\ are unambiguous, i.e., for every \owlifier\ block
(or set of blocks in the case of sufficient blocks) there is a
well-defined set of corresponding OWL-DL axioms. This property is
important because it implies that reasoning can be performed over
ontologies defined in \owlifier\ using standard OWL-DL reasoners.  We
use this capability in our current \owlifier\ implementation
(described further in \secref{sec:implementation}) to verify
ontologies defined using \owlifier\ and report possible errors to
users.

In general, new axioms are inferred from an \owlifier\ ontology
primarily from the use of synonym blocks, sufficient blocks, and
transitive blocks (whose inferences are described in the previous
section). For instance, let $A$, $B$, and $C$ be classes and $P$ be an
object property. From an axiom $A \equiv B$ generated from a synonym
block, and an axiom $B \sqsubseteq \exists P.C$ generated from a
relationship block, the axiom $A \sqsubseteq \exists P.B$ would be
inferred. Thus, the axioms of a particular class are ``inherited'' by
all of the synonyms of the class. Similarly, from an axiom $A \equiv
\exists P. C$ generated from a sufficent block, and an axiom $B
\sqsubseteq \exists P.C$ generated from a relationship block, the axiom
$B \sqsubseteq A$ would be inferred. For both the case of synonym and
sufficient blocks, the use of equivalence permits a number of
additional inferences to be made via an OWL-DL reasoner.

As described in \cite{rector04:_owl_pizzas} additional reasoning can
occur within OWL-DL ontologies when domain and range axioms are
providedq (as well as, e.g., property closure axioms).  We explicitly
do not consider these constraints in the current version of \owlifier\
because they also often result in modeling errors for inexperienced
OWL-DL users \cite{rector04:_owl_pizzas}. Instead, we adopt the more
traditional description-logic approaches that do not have explicit
domain and range axioms.  Although not currently supported in
\owlifier, domain and range constraints could be inferred from given
relationship blocks as well as property closure axioms.

@@

To help simplify the creation of class hierarchies and property
sequences (including transitive, cardinality, and sufficient blocks),
we allow for ``sparse'' blocks that inherit missing information from
their closest proceeding block. For instance the following entity
blocks
\begin{tabbing}
  ~~\textsf{entity} \= \textsf{EcologicalHabitat} \= \textsf{Estuary} \= \kill
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Estuary} \> 
  \textsf{Bay}\\
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Estuary} \>
  \textsf{Fjord} \\
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Marsh} \>
  \textsf{TidalMarsh} \\ 
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Marsh} \>
  \textsf{SaltMarsh}
\end{tabbing}
can be equivalently represented in \owlifier\ using the following
sparse blocks.
\begin{tabbing}
  ~~\textsf{entity} \= \textsf{EcologicalHabitat} \= \textsf{Estuary} \= \kill
  ~~\textsf{entity} \> \textsf{EcologicalHabitat} \> \textsf{Estuary} \> 
  \textsf{Bay}\\
  \> \> \> \textsf{Fjord} \\
  \> \> \textsf{Marsh} \> \textsf{TidalMarsh} \\ 
  \> \> \> \textsf{SaltMarsh}
\end{tabbing}
In general, the use of sparse blocks simplifies ontology creation by
not requiring users to enter every (redundant, leading) field to be
explicitly given, which in turn can simplify the overall layout of the
ontology within a spreadsheet.  Additionaly, \owlifier\ does not place
constraints on the order of blocks within a spreadsheet. Classes also
do not need to explicitly be defined within an entity block, e.g.,
classes without corresponding entity blocks can be introduced simply
through synonym and relationship blocks. This typically occurs when a
particular class does not participate as a subclass or superclass of
another class in the current ontology. Overall, one of our goals is
for \owlifier\ to allow minimal information to be entered by limitting
redundancy as well as by supporting a range of inferences, while at
the same time reducing the number of errors commonly made in ontology
development by non-experts.

\section{The \Owlifier Application}
\label{sec:implementation}

\figref{fig:owlifier} shows the general architecture of the \owlifier\
application. A scientist or researcher first creates a spreadsheet
containing a set of ontology definitions using \owlifier\ conventions.
The spreadsheet is then exported as a plain-text file containing the
\owlifier\ table, e.g., using a CSV or tab-delimited format. The
\owlifier\ table is then sent to the \owlifier\ application, which
performs a number of steps that include: (i) parsing the file; (ii)
generating the corresponding OWL-DL representation (e.g., in the
current implementation, the Jena API is used \cite{carroll04:_jena});
(iii) validating the ontology and ensuring it is consistent via an
OWL-DL reasoner (e.g., the current implementation uses Pellet
\cite{sirin07:_pellet}); and (iv) outputing the corresponding OWL-DL
file. The resulting OWL file can then be refined and extended via
existing OWL tools (e.g., Protege or SWOOP). It is possible that the
refined ontology is converted back to a corresponding \owlifier\
representation (shown using dashed lines in
\figref{fig:owlifier}). While simple to perform, this last conversion
is not information preserving since not all OWL-DL constructs are
represented through \owlifier\ blocks.

The current implementation of the \owlifier\ application is written in
Java and supports a subset of the blocks defined in
\secref{sec:owlifier}. In particular, we are currently extending
\owlifier\ to support cardinality and sufficient blocks as well as to
perform the above conversion from OWL-DL files to corresponding
\owlifier\ blocks. As noted above, the current implementation,
although it does not support all of the blocks defined here, is being
used within a project focused on supporting trait data and many of the
extensions outlined here are based on these experiences. In addition,
\owlifier\ has been used to construct term hierarchies from large sets
of keywords (harvested from existing metadata documents) as well as to
define articulations into the observation ontology defined in
\cite{madin07:_ontol_for_descr_and_synth}.

\begin{figure}
  \centering
  \includegraphics[scale=.5]{architecture.pdf}
  \caption{The basic \owlifier\ application and relation to other
    technologies.}
  \label{fig:owlifier}
\end{figure}

@@

We describe a new To address barriers for ontology development and adoption within
ecology, we have presented a new approach that allows the \owlifier,
which consists of a set orelated work: similar to dave's work on
``global taxonomic constraints'', which assume certain structures of
hierarchies, e.g., disjoint siblings. protege approach/plugin
